---
output: 
  html_document: 
    keep_md: yes
---

```{r echo=FALSE}
library = function (...) suppressMessages(base::library(...))
library("dplyr")
library("ggplot2")
library("knitr")

#Load timezones
timezones_df = read.table("data/airport_timezones.csv", sep = ",", 
                          header = TRUE, colClasses = c("character", "character"), stringsAsFactors = FALSE)

#Load data
data = read.table("data/ryanair_checkin_data.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE) %>%
  dplyr::mutate(checkin_datetime =  paste(checkin_date, checkin_time)) %>%
  dplyr::mutate(departure_datetime = paste(departure_date, departure_time)) %>%
  dplyr::left_join(timezones_df, by = "from") %>%
  dplyr::tbl_df()

#Caclulate difference in days
checkin = strptime(data$checkin_datetime, "%d/%m/%Y %H:%M", tz = "GMT")
departure = strptime(paste(data$departure_datetime, data$timezone), "%d/%m/%Y %H:%M:%S %z", tz = "GMT")
data = dplyr::mutate(data, days_before_departure = as.numeric(departure-checkin))

#Extract row letters and numbers
data_seat_classes = dplyr::mutate(data, seat_letter = stringr::str_extract(data$seat_number, '\\D')) %>%
  dplyr::mutate(seat_row = as.numeric(stringr::str_extract(data$seat_number, '\\d+'))) %>%
  dplyr::mutate(seat_is_middle = ifelse(seat_letter == "B" | seat_letter == "E", TRUE, FALSE)) %>%
  dplyr::mutate(checkin_first_day = ifelse(days_before_departure >= 6, TRUE, FALSE)) %>%
  dplyr::mutate(checkin_text = ifelse(checkin_first_day, "Within 24 hours", "After 24 hours"))

#Calculate the frequency of getting middle seats
freq_table = dplyr::group_by(data_seat_classes, checkin_text) %>% 
  summarise(middle_seat_count = length(which(seat_is_middle)), total_count = length(seat_is_middle)) %>%
  dplyr::select(checkin_text, total_count, middle_seat_count) %>%
  dplyr::mutate(frequency = middle_seat_count/total_count) %>%
  as.data.frame()
colnames(freq_table) = c("Check-in time", "Number of flights", "Number of middle seats", "Fraction")
```


# How to avoid middle seats on Ryanair flights?

I'm a frequent flier with Ryanair and I like their assigned seats policy, but I really don't like sitting on the middle seat. Their website makes it seem that the seats are assigned *randomly* and the only way to avoid certain undesired seats is to pay the 6 GBP/EUR extra fee and choose your own seat. But maybe their seat allocation is not actually completely random and it might be possible to predict which seat you are likely to get? 

Some time ago, it seemed to me that when I checked in really early I was less likely to sit in the middle compared to checking in later. To check if this was really the case, I compared the check-in times to the deparure times for all 46 Ryanair flights that I have taken during the last two years. Indeed, if I checked in during the first 24 hours, I ended up sitting in the middle seat only on 1/29 (3.5%) flights whereas this increased to 8/17 (47%) when I check in later:

```{r echo=FALSE}
kable(freq_table, format = "markdown")
```
So there you have it, if you do not want to sit in the middle then checking in early does indeed look very promising.

But are there any downsides to early check-in? Yes, you are likely to be one of the last person to get off the plane when it lands. That's because the seat allocation tends to start from the rows that are in the middle of the plane. The follwing figure shows how check-ins done within the first 24 hours almost always give you a seat in rows between 15-22 (exluding emergency exit rows 16 and 17). On the otherhand, for check-ins done later there is no strong row bias and you are more likely to sit eihter towards the fron end or back end of the plane.

```{r echo=FALSE}
ggplot(data_seat_classes, aes(x = as.factor(seat_row))) + geom_histogram() + 
  facet_wrap(~checkin_text) + 
  xlab("Seat row") + 
  ylab("Number of flights") + 
  theme_bw()
```

##Technical details
* **Where did you get the check-in times from?** Whenever I check in for a flight, I always download the PDF copy of the boarding pass and save it onto a folder on my computer. This means that all of these files will have last modification date attached to them. The last modification date corresponds almost perfectly to the time that I checked in for the flight.
* **Is this analysis reproducible? ** Yes! All of the raw data is the the data/ subfolder of this repository and all of the neccessary code is in the README.Rmd document.
* **Can I condtribute? ** Of course! If you've got lot's of Ryanair boarding pass PDFs lying around that have the correct check-in times attached to them, then add them to the 'data/ryanair_chedkin_data.csv' and send them to me either via e-mail (kauralasoo [at] gmail.com) or as a github pull request. If you want to do additional analysis on this data then feel free to fork this repository.



